Structuring College Access: The Market Segment Model and College Board Geo-markets

### 2024-10-18 text/old_spring2024/front_end.qmd
revising front end

In 1983, Zemsky and Oedel authored *The Structure of College Choice*, published by *The College Board.* This book develops the Market Segment Model and introduces local "community-based enrollment markets" (p.14) -- later called "geomarkets" -- which are the basis for the College Board market Enrollment Planning Service (EPS), a market research product. @RN4982[pp., 9-10] write that the Market Segment Model sought to quantify the knowledge that college admissions officers have about student demand:

@@

In 19XX, geomarkets were incorporated into the College Board Student Search Service student list product, which College Board created in 1972 [@belkin2019-studata]. Student lists contain the contact information of prospective students, the basic input for recruiting campaigns that target individual prospects using mail, email, and targeted social media. RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 20XX) and control which prospect profiles through purchase by selecting search filters, such as high school graduation year, SAT score, AP test score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular geomarkets.

@@

However, prior research has not examined the role of third-party market research in reproducing socioeconomic and racial inequality. This manuscript analyzes the Market Segment Model and College Board Geomarkets as a case study of quantification. We develop arguments from the broader literature on quantification [e.g.,] [@RN5003; @RN5005; @RN5007], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class= [@RN4982]. Homophily is the idea that actors that share common characteristics are likely to form connections with one another. The Market Segment Model argues that homophily is the organizing principle of student demand and competition between colleges, that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Subsequently, the assumption of homophily is programmed into the EPS and Student Service Service products that colleges utilize to decide where to focus recruiting energy.

@@

```{=html}
<!--The enrollment management industry emphasizes market research.
Market research is the process of producing information that informs decisions about product design or how to find customers.   -->
```

The enrollment management industry puts great stock in market research. @RN4986[p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].^[@RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."] However, most market research in the contemporary enrollment management industry is produced by third-party organizations. For example, Ruffalo-Noel Levitz and EAB both produce prodigious amounts of market research, some publicly available as a means of business development [e.g., @RN4741; @RN4728, but most only available to clients. EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT.

Despite being central to enrollment management for at least 40 years [@RN4986], the role of third-party market research has been ignored by scholarship on college access. Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935;@RN1814]. However, market research informs where colleges look for students and which prospective students to target. These recruiting interventions intervene on student demand [@RN4324; @cottom2017lower], thereby structuring college access pathways in ways that could amplify or diminish historic inequality in college access. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals [@RN4986]. This article offers one story about the quiet revolution that transformed college access into a sophisticated, supply-led enterprise. We argue that market research structures who colleges recruit, in ways that reinforce existing social stratification. The case of Market Segment Model and College Board geomarkets shows how market research -- and its commodification into recruiting products -- is a mechanism for social reproduction. 
[LAST TWO SENTENCES ARE REDUNDANT BUT BUT ARE TRYING TO DO SLIGHTLY DIFFERENT THINGS AND I LIKE THEM BOTH]

@@

```{=html}
<!--
CUT GENERAL PARAGRAPH ON REPRODUCTION

Within the sociology of education, @RN2377 describes education as a sieve, allocating people to opportunities in socially stratified, capitalist societies. This role highlights "Max Weber's core insight that education has a dual character" [@RN2377, p. 129], on one hand, serving social mobility by redistributing access to privileged positions and, on the other, serving social reproduction by granting privileged families disproportionate access to the educational institutions that confer pathways to privileged positions. Similarly, @RN4873 describe schools and colleges as "sorting machines" that incorporate external categories (e.g., race, class), create categories (e.g., instructional tracks), and allocate students to categories. Initial category allocation affects subsequent allocation, such that school assignments and instructional track assignments affect college access, which affects employment assignments. "In evaluating schools as inequality regimes" @RN4873 \[p. 322\] argue, "the relevant question is not whether schools produce inequality; rather, the relevant questions are, when and where do schools produce inequality, \[and\] how do they do so." As @RN2377 show, whereas early research on social stratification sought to identify patterns of inequality [e.g., @RN4873; @RN290; @RN1268], the sociology of education is now more concerned with identifying the mechanisms underlying these patterns. 
-->
```
We situate our study amidst scholarship on within scholarship on quantification as a mechanism that reproduces inequality. @RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402) and introduce reacivity, discipline, and authority as three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Discipline is the idea that quantification causes actors to react in a particular way. Authority is the idea that quantification changes decision-making power, often weakening the discretion of local decision-makers.

@@

```{=html}
<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->
```
In the interdisciplinary literature on quantification [@RN5008], the discussions of correlation and homophily by @RN4975 introduce concepts that are salient to the analysis of market research. Chapter 1 of @RN4975 is *Correlating Eugenics*. Correlation measures the extent to which two or more variables move together. The Pearson Correlation Coefficient was developed by the 19th Century statisticians and "biometric eugenicists" Francis Galton and his student Karl Pearson [@RN4975]. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared in his 2008 editorial 'The End of Theory,' big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'" Predictive analytics are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes.

@@

Chapter 2 of @RN4975, *Homophily, or the Swarming of the Segregated Neighborhood*, argues that homophily often underlies the use of correlation to predict future behavior. Homophily means that actors who share common characteristics are likely to form connections with one another. The ubiquitous colloquialism is "birds of a feather flock together." Homophily is a core assumption for network science [@mcpherson2001birds], in which actors (e.g., high school students) are linked to one another through direct and indirect network ties (e.g., submit SAT scores to the same college). Because network science often draws from rational choice theory, it assumes that homophily is the result of voluntary action by individuals, thereby "eras\[ing\] historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that underlie behavior consistent with homophily. @RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. In commercial social networks, such as Facebook, Twitter/X, and TikTok, homophily is programmed into the algorithms that create connections between users.[^4]

[^4]: Thus, @RN4975 [p. 82] writes, "echo chambers are not unfortunate errors, but deliberate goals" because "homophily is used to create agitated clusters of individuals whose angry similarity and overwhelming attrication to their common object of hatred both repel them from one another and glue them together."

@@

 <!--Historically, College Board partnered with Members to develop projects that later became revenue-generating products for College Board.-->
As a member organization, several College Board products developed from joint collaborations with administrators or professors from member institutions. During the 1970s, under the leadership of Larry Litten, then director of institutional research, Carleton College pionered applying market research to student recruiting. In the late 1970s, College Board provided Carleton College with funding and data for a six-year project that analyzed college choice decisions by households in six metropolitan areas. This project was the basis for the landmark book *Applying Market Research in College Admissions* [@RN4986], published by College Board, and was the basis for the College Board Admitted Student Questionnaire (ASQ), a revenue-generating product that provides colleges with information about admitted students -- those who enrolled and who did not enroll. Litten would later become Associate Director of the Consortium of Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s.

In 1978, Robert Zemsky, a University of Pennsylvania professor, was asked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. To answer these questions, Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE). As the project project became more ambitious, the researchers reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest demographic unit used in tehse analyses was the state' [@RN4982, p. x]. In 1979, College Board began providing data and funding for what became the *Comprehensive Undergraduate Enrollment Planning Project (CUEPP)*.^[say that Pennsylvania Department of Education was also a partner; see page x of Zemsky] @RN4982[p. 4] write that,

@@

**Creating the Market Segment Model.** Considering the @RN5007 \[p. 402\] definition of quantification as "the production and communication of numbers," the Market Segment Model is an effort to quantify and communicate the knowledge that admissions officers have about student demand. @RN4982[p. 4] state that "our initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools." They created  "three types of boundaries" [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. Next, "we divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" \[in New England, Middle States, and the South\].

@@

The two primary outputs of the market segment model are the (1) Market Segment Profile and (2) the Institutional Profile. The Market Segment Profile, which was produced separately for each of the 143 geo-markets, provides information about the number and characteristics of students in each market segment. @tbl-market-segment-characteristics reproduces a partial, simplified version of the bottom panel of @RN4982 [Figure 2.1] which represents the Market Segment Profile for "Connecticut Market 3: Fairfield County." @tbl-market-segment-characteristics has separate columns for each market segment (local, in-state, regional, national) and rows show the number of test-takers and characteristics of test-takers. For example, there were 550 "local" students in Fairfield county and these students submitted SAT scores to 2.8 postsecondary institutions on average. By contrast, there were 1,664 "regional" students who submitted test scores to 4.8 institutions on average. For local students, 11.2% had family income greater than \$35,000 and 9% had both parents with a BA. For regional students, 41.9% had family income greater than \$35,000 and 34.0% had both parents with a BA. Each Market Segment Profile also present information about the institutions that students from each segment sent scores to.[^5]

[^5]: @tbl-market-segment-institutions recreates the top panel of @RN4982 [Figure 2.1] for the Fairfield County, CT local market. For example, of the 550 local students, 277 (50.4%) sent scores to institution #1, which was a private master's granting institution. Of the 1,199 in-state students, 757 (63.1) sent scores to institution #1, a public doctoral granting institution, and 515 (43.0%) sent scores to institution #2, a public master's granting institution.

@@

@RN4982 summarize their correlational analyses using authoritative language. @RN4982[pp. 34-35] asserts, "it is hard to overemphasize the importance of this statistical pattern. Social data seldom line up as we expect \ldots The information derived from the Market Segment Model, however, is a remarkably ordered set of data, consistent in its relationships, that reflects the basic social and economic patterning associated with the structure of college choice." With respect to family income, @RN4982[p. 33] state that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982[p. 33] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time." Further analyses reveal that the SAT score is the most important predictor of score-sending behavior, followed by parental education, family income, and educational aspirations. Considering prior research showing that SAT scores are substantially a function of social origin [@RN1273], the Market Segment Model implicitly and explicitly states that student demand for higher education is mostly a function of social origin. @RN4982 also find that the SAT score, parental education, and family income predict student score-sending behavior at the local geo-market level.

However, geomarkets differ in the relative abundance of students with particular socioeconomic characteristics and this has practical implications for recruiting. @RN4982 recommend that colleges target geo-markets with desirable compositions of socioeconomic characteristics and student market segment. @RN4982[p. 44] describes a hypothetical college that wants to target regional and national students in New England and is considering whether to expend limited recruiting efforts in the Boston geo-market -- which was relatively low-income in 1980 -- or in more affluent geo-markets nearby:

@@

```{=html}
<!--
@RN5019 [2005]
> The marketing software that pinpoints the schools and geomarkets where your best prospects are most likely to be found. With the click of a mouse, EPS provides you with comprehensive reports on your markets, your position in those markets, and your competition. Focus your valuable time and resources on the right prospects

Software documentation by @RN5017 states that "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas, mostly in the United States. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."
-->
```
Whereas @RN4982[Appendices B1-B3] identified 143 geo-markets covering the New England, Middle States, and South region, EPS created geo-markets for the remaining U.S. states, with 304 geo-markets in total. @RN4988 shows the College Board geo-markets circa 2023. Interestingly, the geo-markets for New England, Middle States, and the South remain nearly identical to those developed by @RN4982. Concrete information about how EPS created the remaining geomarkets is elusive. Documentation and promotional material suggests that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.[^7]

[^7]: @RN5016 states that "geomarkets are areas within a state that represent a further segmentation of a population. Students from California don't all share the same college-going behaviors. We have accounted for this variance by segmenting the 50 states into 304 geomarkets to provide further insight into student behaviors within particular areas of individual states."

Although how we lack clarity about how geomarkets were created, we expect substantial socioeconomic inequality between geomarkets in large metropolitan areas. Given the extent of class- and race-based residential segregation in the U.S., it would be surprising to not observe such inequality. Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982[pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:

@@

**Homophily and Utilization**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] shows that homophily is a core tenet of the Market Segment Model. The authors conduct a social network analysis that defines two institutions as being in competition with one another if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa.[^8] @RN4982[p. 42] sate, "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions."[^9] Subsequent analyses investigate the tuition price and socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. @RN4982[p. 72] describe these patterns as a natural process of homophily in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

[^8]: We can think of students sending SAT scores to colleges as a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). Next, the authors develop "Tinker Toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market.

[^9]: For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."

>>>>

In 1983, Robert Zemsky and Penney Oedel authored *The Structure of College Choice*, published by *The College Board.* This book develops the Market Segment Model and introduces local "community-based enrollment markets" (p.14) -- later called "geomarkets" -- which are the basis for the College Board market Enrollment Planning Service (EPS), a market research product. @RN4982 [pp., 9-10] write that the Market Segment Model sought to quantify the knowledge that college admissions officers have about the geography of student demand:

@@

In 19XX, geomarkets were incorporated into the College Board Student Search Service student list product, which College Board created in 1972 [@belkin2019-studata]. Student lists contain the contact information of prospective students, the basic input for recruiting campaigns that target individual prospects using mail, email, and targeted social media. @RN4895 reported that 87% of private and 86% of public four-year institutions purchase student lists. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50 in 2021) and control which prospect profiles through purchase by selecting search filters, such as high school graduation year, SAT score, AP test score, state, etc. Geomarket filters enable colleges to include/exclude prospects from particular geomarkets.

@@

However, prior research has not examined the role of third-party market research in reproducing socioeconomic and racial inequality. This manuscript analyzes the Market Segment Model and College Board Geomarkets as a case study of quantification. We develop arguments from the broader literature on quantification \[e.g.,\] [@RN5003; @RN5005; @RN5007], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality. The Market Segment Model analyzes SAT score-sending data from 1980 and concludes that student demand for higher education is primarily a function of social class= [@RN4982]. Homophily is the idea that actors that share common characteristics are likely to form connections with one another. The Market Segment Model argues that homophily is the organizing principle of student demand and competition between colleges, that "the hierarchical structure of collegiate competition largely reflects the stratified social and economic dimensions of the communities from which colleges draw their students" [@RN4982, p. 72]. Subsequently, the assumption of homophily is programmed into the EPS and Student Service Service products that colleges utilize to decide where to focus recruiting energy.

@@

<!--The enrollment management industry emphasizes market research.
Market research is the process of producing information that informs decisions about product design or how to find customers.   -->

The enrollment management industry puts great stock in market research. @RN4986 [p. 17] define market research as all "activities intended to increase understanding of a market and consumer behavior in that market." At the most basic level, in-house market research is a function of being intentional about record-keeping and using those records [@RN3519].[^1] However, most market research in the contemporary enrollment management industry is produced by third-party organizations. For example, Ruffalo-Noel Levitz and EAB both produce prodigious amounts of market research, some publicly available as a means of business development \[e.g., @RN4741; @RN4728, but most only available to clients. EPS remains the market research arm of College Board, while Encoura Eduventures is the market research arm of ACT.

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."

Despite being central to enrollment management for at least 40 years [@RN4986], the role of third-party market research has been ignored by scholarship on college access. Scholarship on college access often takes a demand-side perspective, whereby household differences in economic and cultural capital yield stratification in student preferences and the ability to gain admission [e.g., @RN4935; @RN1814]. However, market research informs where colleges look for students and which prospective students to target. These recruiting interventions intervene on student demand [@RN4324; @cottom2017lower], thereby structuring college access pathways in ways that could amplify or diminish historic inequality in college access. The College Board played a pivotal role in transforming market research about enrollment from something done by individual universities to something done by third-party professionals [@RN4986]. This article offers one story about the quiet revolution that transformed college access into a sophisticated, supply-led enterprise. We argue that market research structures who colleges recruit, in ways that reinforce existing social stratification. The case of Market Segment Model and College Board geomarkets shows how market research -- and its commodification into recruiting products -- is a mechanism for social reproduction. \[LAST TWO SENTENCES ARE REDUNDANT BUT BUT ARE TRYING TO DO SLIGHTLY DIFFERENT THINGS AND I LIKE THEM BOTH\]

@@

<!--
CUT GENERAL PARAGRAPH ON REPRODUCTION

Within the sociology of education, @RN2377 describes education as a sieve, allocating people to opportunities in socially stratified, capitalist societies. This role highlights "Max Weber's core insight that education has a dual character" [@RN2377, p. 129], on one hand, serving social mobility by redistributing access to privileged positions and, on the other, serving social reproduction by granting privileged families disproportionate access to the educational institutions that confer pathways to privileged positions. Similarly, @RN4873 describe schools and colleges as "sorting machines" that incorporate external categories (e.g., race, class), create categories (e.g., instructional tracks), and allocate students to categories. Initial category allocation affects subsequent allocation, such that school assignments and instructional track assignments affect college access, which affects employment assignments. "In evaluating schools as inequality regimes" @RN4873 \[p. 322\] argue, "the relevant question is not whether schools produce inequality; rather, the relevant questions are, when and where do schools produce inequality, \[and\] how do they do so." As @RN2377 show, whereas early research on social stratification sought to identify patterns of inequality [e.g., @RN4873; @RN290; @RN1268], the sociology of education is now more concerned with identifying the mechanisms underlying these patterns. 
-->

We situate our study amidst scholarship on within scholarship on quantification as a mechanism that reproduces inequality. @RN5007 [p. 402] define quantification as "the production and communication of numbers" (p. 402) and introduce reacivity, discipline, and authority as three interrelated themes that describe the effects of quantification. Reactivity is the idea that salient quantitative measures cause people and organizations to change their behavior. Discipline is the idea that quantification causes actors to react in a particular way. Authority is the idea that quantification changes decision-making power, often weakening the discretion of local decision-makers.

@@

<!-- THIS IS THE CUT TRANSITION PARAGRAPH
In the broader quantification literature, critical data studies [e.g., @RN4975; @RN4849; @RN4772] and sociological analyses of technology and inequality [e.g., @RN4778; @RN4786; @RN4877; @RN4810] introduce concepts salient to the analysis of market research. Our focus is on "selection devices," which utilize some combination of inputs and discretion to categorize places, people, and organizations for the purpose of allocating opportunities or interventions [@RN4778; @RN5012].
-->

In the interdisciplinary literature on quantification [@RN5008], the discussions of correlation and homophily by @RN4975 introduce concepts that are salient to the analysis of market research. Chapter 1 of @RN4975 is *Correlating Eugenics*. Correlation measures the extent to which two or more variables move together. The Pearson Correlation Coefficient was developed by the 19th Century statisticians and "biometric eugenicists" Francis Galton and his student Karl Pearson [@RN4975]. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared in his 2008 editorial 'The End of Theory,' big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'" Predictive analytics are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes.

@@

Chapter 2 of @RN4975, *Homophily, or the Swarming of the Segregated Neighborhood*, argues that homophily often underlies the use of correlation to predict future behavior. Homophily means that actors who share common characteristics are likely to form connections with one another. The ubiquitous colloquialism is "birds of a feather flock together." Homophily is a core assumption for network science [@mcpherson2001birds], in which actors (e.g., high school students) are linked to one another through direct and indirect network ties (e.g., submit SAT scores to the same college). Because network science often draws from rational choice theory, it assumes that homophily is the result of voluntary action by individuals, thereby "eras\[ing\] historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that underlie behavior consistent with homophily. @RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. In commercial social networks, such as Facebook, Twitter/X, and TikTok, homophily is programmed into the algorithms that create connections between users.[^2]

[^2]: Thus, @RN4975 [p. 82] writes, "echo chambers are not unfortunate errors, but deliberate goals" because "homophily is used to create agitated clusters of individuals whose angry similarity and overwhelming attrication to their common object of hatred both repel them from one another and glue them together."

@@

<!--Historically, College Board partnered with Members to develop projects that later became revenue-generating products for College Board.-->

As a member organization, several College Board products developed from joint collaborations with administrators or professors from member institutions. During the 1970s, under the leadership of Larry Litten, then director of institutional research, Carleton College pionered applying market research to student recruiting. In the late 1970s, College Board provided Carleton College with funding and data for a six-year project that analyzed college choice decisions by households in six metropolitan areas. This project was the basis for the landmark book *Applying Market Research in College Admissions* [@RN4986], published by College Board, and was the basis for the College Board Admitted Student Questionnaire (ASQ), a revenue-generating product that provides colleges with information about admitted students -- those who enrolled and who did not enroll. Litten would later become Associate Director of the Consortium of Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s.

In 1978, Robert Zemsky, a University of Pennsylvania professor, was asked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. To answer these questions, Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE). As the project project became more ambitious, the researchers reached out to College Board because "we needed a database that described most institutions and most students\ldots Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest demographic unit used in tehse analyses was the state' [@RN4982, p. x]. In 1979, College Board began providing data and funding for what became the *Comprehensive Undergraduate Enrollment Planning Project (CUEPP)*.[^3] @RN4982 [p. 4] write that,

[^3]: say that Pennsylvania Department of Education was also a partner; see page x of Zemsky

@@

**Creating the Market Segment Model.** Considering the @RN5007 \[p. 402\] definition of quantification as "the production and communication of numbers," the Market Segment Model is an effort to quantify and communicate the knowledge that admissions officers have about student demand. @RN4982 [p. 4] state that "our initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools." They created "three types of boundaries" [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. Next, "we divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" \[in New England, Middle States, and the South\].

@@

The two primary outputs of the market segment model are the (1) Market Segment Profile and (2) the Institutional Profile. The Market Segment Profile, which was produced separately for each of the 143 geo-markets, provides information about the number and characteristics of students in each market segment. @tbl-market-segment-characteristics reproduces a partial, simplified version of the bottom panel of @RN4982 [Figure 2.1] which represents the Market Segment Profile for "Connecticut Market 3: Fairfield County." @tbl-market-segment-characteristics has separate columns for each market segment (local, in-state, regional, national) and rows show the number of test-takers and characteristics of test-takers. For example, there were 550 "local" students in Fairfield county and these students submitted SAT scores to 2.8 postsecondary institutions on average. By contrast, there were 1,664 "regional" students who submitted test scores to 4.8 institutions on average. For local students, 11.2% had family income greater than \$35,000 and 9% had both parents with a BA. For regional students, 41.9% had family income greater than \$35,000 and 34.0% had both parents with a BA. Each Market Segment Profile also present information about the institutions that students from each segment sent scores to.[^4]

[^4]: @tbl-market-segment-institutions recreates the top panel of @RN4982 [Figure 2.1] for the Fairfield County, CT local market. For example, of the 550 local students, 277 (50.4%) sent scores to institution #1, which was a private master's granting institution. Of the 1,199 in-state students, 757 (63.1) sent scores to institution #1, a public doctoral granting institution, and 515 (43.0%) sent scores to institution #2, a public master's granting institution.

@@

@RN4982 summarize their correlational analyses using authoritative language. @RN4982 \[pp. 34-35\] asserts, "it is hard to overemphasize the importance of this statistical pattern. Social data seldom line up as we expect \ldots The information derived from the Market Segment Model, however, is a remarkably ordered set of data, consistent in its relationships, that reflects the basic social and economic patterning associated with the structure of college choice." With respect to family income, @RN4982 [p. 33] state that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982 [p. 33] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time." Further analyses reveal that the SAT score is the most important predictor of score-sending behavior, followed by parental education, family income, and educational aspirations. Considering prior research showing that SAT scores are substantially a function of social origin [@RN1273], the Market Segment Model implicitly and explicitly states that student demand for higher education is mostly a function of social origin. @RN4982 also find that the SAT score, parental education, and family income predict student score-sending behavior at the local geo-market level.

However, geomarkets differ in the relative abundance of students with particular socioeconomic characteristics and this has practical implications for recruiting. @RN4982 recommend that colleges target geo-markets with desirable compositions of socioeconomic characteristics and student market segment. @RN4982 [p. 44] describes a hypothetical college that wants to target regional and national students in New England and is considering whether to expend limited recruiting efforts in the Boston geo-market -- which was relatively low-income in 1980 -- or in more affluent geo-markets nearby:

@@

<!--
@RN5019 [2005]
> The marketing software that pinpoints the schools and geomarkets where your best prospects are most likely to be found. With the click of a mouse, EPS provides you with comprehensive reports on your markets, your position in those markets, and your competition. Focus your valuable time and resources on the right prospects

Software documentation by @RN5017 states that "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas, mostly in the United States. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."
-->

Whereas @RN4982 [Appendices B1-B3] identified 143 geo-markets covering the New England, Middle States, and South region, EPS created geo-markets for the remaining U.S. states, with 304 geo-markets in total. @RN4988 shows the College Board geo-markets circa 2023. Interestingly, the geo-markets for New England, Middle States, and the South remain nearly identical to those developed by @RN4982. Concrete information about how EPS created the remaining geomarkets is elusive. Documentation and promotional material suggests that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.[^5]

[^5]: @RN5016 states that "geomarkets are areas within a state that represent a further segmentation of a population. Students from California don't all share the same college-going behaviors. We have accounted for this variance by segmenting the 50 states into 304 geomarkets to provide further insight into student behaviors within particular areas of individual states."

Although how we lack clarity about how geomarkets were created, we expect substantial socioeconomic inequality between geomarkets in large metropolitan areas. Given the extent of class- and race-based residential segregation in the U.S., it would be surprising to not observe such inequality. Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982 [pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:

@@

**Homophily and Utilization**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] shows that homophily is a core tenet of the Market Segment Model. The authors conduct a social network analysis that defines two institutions as being in competition with one another if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa.[^6] @RN4982 [p. 42] sate, "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions."[^7] Subsequent analyses investigate the tuition price and socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. @RN4982 [p. 72] describe these patterns as a natural process of homophily in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

[^6]: We can think of students sending SAT scores to colleges as a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). Next, the authors develop "Tinker Toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market.

[^7]: For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."
